### 2019-04-29 reports/article.tex
dropped sls clc comparison

\section*{Abstract}

Urbanization is currently a global phenomenon that has become the most important form of landscape change and is increasingly affecting biodiversity and ecosystem functions.
In order to evaluate the impacts of urbanization and inform urban planning, it is important to understand the spatiotemporal patterns of land use change associated to urbanization.
This paper exploits three different frameworks, namely landscape metrics, urban growth modes and fractal analysis to characterize the spatiotemporal patterns of urbanization of the Swiss urban agglomerations of Zurich, Bern and Lausanne. % from 1980 to 2016
The land use inventory provided by the Swiss Federal Statistical Office were used to assemble four temporal snapshots from 1980 to 2016 at the extent of the urban agglomerations.
The time series of landscape metrics generally supports the diffusion and coalescence hypothesis, yet many remaining natural patches hinder the coalescence of urban patches.
On the other hand, an analysis of the urban growth modes suggest that leapfrog development occurs at all periods, which contributes to an increasing fragmentation of natural patches and maintains the fractal configuration of the landscape.
The discussion reviews potential explanations for the observed landscape changes, and concludes with some planning implications.

@@

Given that a significant part of the cross-border urban agglomerations of Basel and Geneva lie beyond the Swiss boundaries \cite{sfso2014espace}, in order to ensure coherence of the land use/land cover data (see the section below), this study comprises only three of the five largest Swiss urban agglomerations, namely Zurich, Bern and Lausanne \cite{sfso2018city}.
As shown in \autoref{figures/population_change.pdf}, the three cities have undergone important population growth over the last 30 years, especially during the most recent years and at the agglomeration extent.
With a total population over 1.3 million and land area of 1305 km$^2$ (1038 hab/km$^2$), Zurich is the largest Swiss urban agglomeration. As a leading global city and one of the world's largest financial centers, Zurich has the country's largest airport and railway station, and also hosts the largest Swiss universities and higher education institutions. 
% ranking among the most liveable cities in the world.
Bern is the capital of Switzerland and fourth most populous urban agglomeration in Switzerland, with a total population of 410000 inhabitants and occupying a land area of 783 km$^2$ (531 hab/km$^2$). As the fifth largest Swiss urban agglomeration and the second most important student and research center after Zurich, Lausanne has a total population of 409000 inhabitants over a land area of 773 km$^2$ (537 hab/km$^2$). Given its larger population growth rate, Lausanne is likely to soon surpass Bern and become the fourth largest urban agglomeration in Switzerland.
% Zurich: 1021859 (1990), 1080728 (2000), 1191058 (2010), 1354140 (2017)
% Bern: 351084 (1990), 349157 (1990), 357668 (2010), 415785 (2017)
% Lausanne: 300280 (1990), 311441 (2000), 339,389 (2010), 415596 (2017)
% Densities: Zurich: 1037 Bern: 531, Lausanne: 537
Overall, the three urban agglomerations are characterized by a pervasive public transportation system and a highly developed economy, with a 85\% of the employment devoted to the tertiary sector.

@@

The two main land use/land cover datasets available for Switzerland are the CORINE land cover (CLC) inventory \cite{heymann1994corine} and the Swiss land statistics (SLS) datasets \cite{sfso2017statistique}.
As part of a pan-European project, the CLC provides an inventory of land cover maps which are standardized among a set of European countries.
The CLC inventory comprises a hierarchical nomenclature of 3 levels, where level 3 comprises 44 land cover classes, and datasets have been released for 1990, 2000, 2006, 2012 and 2018, with a spatial resolution of one hectare per pixel. % \footnote{Although Switzerland was not part of the first CLC inventory for the year 1990, Switzerland published in 1998 a compatible dataset for CLC 1990 \cite{steinmeier2013corine}. By the release of the following CLC dataset in 2000, Switzerland was already a member of the CORINE project, therefore the CLC datasets for 2000, 2006, 2012 and 2018 already include the country.}.
% The SLS land use land cover maps are assembled by the Swiss Federal Statistical Office (SFSO) as part of the production process of the national maps 
On the other hand, the SLS inventory provides maps for the national extent of Switzerland which comprise 72 base categories, and four datasets have been released for 1979/85, 1992/97, 2004/09 and 2013/18\footnote{\label{fn:sls-years}The exact dates of each surveying period 1979/85, 1992/97, 2004/09 and 2013/18 are determined according to the production process of the national maps and vary accross the Swiss territory \cite{sfso2017statistique}}, also with the spatial resolution of one hectare per pixel.

Given their pan-European extent, the CLC inventory would allow the examination of the cross-border urban agglomerations of Geneva and Basel, nevertheless, two main reasons favor the use of the SLS datasets in the context of the present study.
On the one hand, Switzerland was not part of the first CLC inventory for 1990, and although a dataset for the Swiss extent compatible with CLC 1990 was released in 1998 \cite{steinmeier2013corine}, it is only available at a coarser resolution of 250m$\times$250m per pixel. Accordingly, employing CLC at the hectare resolution for Switzerland would be possible over the 2000-2018 period. Instead, SLS provides consistent datasets from 1979 to 2018, therefore allowing a more thorough study of the spatiotemporal patterns of urbanization.
% principle of predominant use, which tends to systematically underpredict the human impacts by completely eliminating human features that occupy modest parts
% interpretation by sampling
On the other hand, while both CLC and SLS rely on computer-aided interpretation of satellite imagery in order to produce the land cover maps, CLC is based on the principle of predominant use, which has been reported to systematically underpredict the human impacts by completely eliminating human features that occupy important yet non-dominant portions of the hectometrical pixel \cite{sfso2017statistique}. In response to such shortcomings, SLS adopted a methodology of interpretation by sampling, which includes special treatment and field verification of pixels where the category attribution is not clear. Since this study intends to explore land use changes related to urban sprawl and fragmentation at the urban fringe, the presence of significant human settlements outside the main urban clusters might not be appreciable within the CLC inventory (see \autoref{figures/sls_clc_comparison.pdf}).

\begin{figure}[!ht]
  \begin{adjustwidth}{-.4\textwidth}{0cm}
    \centering  
    \includegraphics[width=\linewidth]{figures/sls_clc_comparison.pdf}
    \vspace{.5em}
    \caption[Comparison of SLS and CLC]{\label{figures/sls_clc_comparison.pdf}Comparison of the SLS and CLC datasets to the register of buildings and dwellings provided by the Swiss Federal Statistical Office (SFSO). The three maps correspond to the same spatial extent, with the city of Bern at its center.}
  \end{adjustwidth}
\end{figure}

% The preparation of the extracts for the urban agglomerations of Zurich, Bern and Lausanne
According to the considerations reviewed above, the SLS have been used to produce a series of categorical maps for each urban agglomeration and time period.
The boundaries of each urban agglomeration have been adopted from the definitions provided by the Swiss Federal Statistical Office (SFSO) \cite{sfso2014espace}, which comprise multiple municipalities and have been established in consideration of population density, proximity between centers, economic activities and commuting behavior.
% The SLS inventory is managed within a relational database where each entry corresponds to the coordinates of a given pixel and its respective LULC class for each survey period\footref{fn:sls-years}. % which can be downloaded in a comma-separated value text file.
Since the SLS inventory is stored within a relational database where each entry corresponds to the coordinates of a given pixel and its respective land use/land cover classes for each survey period, an open source reusable toolbox to manage, transform and export categorical raster maps has been developed in Python \cite{bosch2019swisslandstats}.
% \cite{bosch2019swisslandstats}.
% The evolution of the urban agglomeration
The spatiotemporal evolution of the urban footprint for the three urban agglomerations is displayed in \autoref{figures/landscape_plots.pdf}.

@@

Firstly, the time series of landscape metrics generally support the diffusion and coalescence hypothesis, with Zurich, the largest Swiss urban agglomeration, exhibiting most characteristics and coalescence, whereas Bern and Lausanne are seemingly undergoing a transition between diffusion and coalescence. Additionally, the trend of SHDI (not explicitly considered in the diffusion and coalescence formulation of Dietzel et al. \cite{dietzel2005spatio}) suggests that the landscape is becoming increasingly diverse in land use. %, which is consistent with the findings of Wu et al. \cite{wu2011quantifying}.

@@

% While the fragmentation of the Swiss spatial planning system 
% The fragmented character of the Swiss spatial planning system is often regarded as \cite{mann2009institutional}. % , messer2017depasser}
% Planning approaches that misappreciate the residential preferences for low-density environments have often resulted 
% Ecosystem services
% The fragmentation of planning authorities is widely regarded as one of the main causes of urban sprawl \cite{carruthers2002fragmentation, mann2009institutional}, and the revision of the Swiss Federal Statue on regional planning of 2013 aims at overcoming such deficiency.
The lack of coordination between of planning authorities is widely regarded as one of the main causes of urban sprawl \cite{carruthers2002fragmentation, mann2009institutional}, and the revision of the Swiss Federal Statue on regional planning of 2013 intends to overcome such shortcoming and define a national growth management strategy based on infill redevelopment and densification.
Nevertheless, growth management policies that misappreciate the residential preferences for low-density environments might inadvertently encourage households to relocate further away from the agglomeration centers, resulting in longer commute times and an overall increase of sprawl at the regional scale \cite{schwanen2004policies, robinson2005twenty}.
Additionally, the interspersion of patches of natural land within urban agglomerations provide valuable ecosystem services to its residents, such as the reduction of air pollution, alleviation of maximum temperatures, absorption of storm water, noise reduction, carbon sequestration, improvement of aesthetic and cultural values as well as the preservation of ecological habitats and biodiversity \cite{bolund1999ecosystem, gomez2013classifying}.
In Switzerland, the lowest supply of such ecosystem services is found in the cantons with highest population density, hence policies encouraging densification should consider the role of natural patches to ensure a sustainable supply of ecosystem services \cite{jaligot2019historical}.
To that end, recent planning approaches based on fractal geometry \cite{yamu2015spatial} might be exploited in order to satisfy the residential demand for green environments while explicitly considering planning objectives such as protecting natural habitats and improving accessibility to urban and natural amenities \cite{bosch2019addressing}.
% Since fractal structures have the key property that their border is over-proportionally lengthened with respect to their area.
% Land use plans should explicitly consider planning goals, fractal planning \cite{bosch2019addressing}

>>>>

\section*{Abstract}
% Urbanization is currently a global phenomenon that has become the most important form of landscape change and is increasingly affecting biodiversity and ecosystem functions. In order to evaluate the impacts of urbanization and inform urban planning, it is important to understand the spatiotemporal patterns of land use change associated to urbanization. This paper exploits three different frameworks, namely landscape metrics, urban growth modes and fractal analysis to characterize the spatiotemporal patterns of urbanization of the Swiss urban agglomerations of Zurich, Bern and Lausanne. The land use inventory provided by the Swiss Federal Statistical Office were used to assemble four temporal snapshots from 1980 to 2016 at the extent of the urban agglomerations. The time series of landscape metrics generally supports the diffusion and coalescence hypothesis, yet many remaining natural patches hinder the coalescence of urban patches.
% On the other hand, an analysis of the urban growth modes suggest that leapfrog development occurs at all periods, which contributes to an increasing fragmentation of natural patches and maintains the fractal configuration of the landscape.
% The discussion reviews potential explanations for the observed landscape changes, and concludes with some planning implications.
Urbanization is currently a global phenomenon that has become the most important form of landscape change and is increasingly affecting biodiversity and ecosystem functions.
In order to evaluate the impacts of urbanization and inform urban planning, it is important to understand the spatiotemporal patterns of land use change associated to urbanization.
This paper exploits three different frameworks, namely landscape metrics, urban growth modes and fractal analysis to characterize the spatiotemporal patterns of urbanization of the Swiss urban agglomerations of Zurich, Bern and Lausanne.
The land use inventory provided by the Swiss Federal Statistical Office were used to assemble four temporal snapshots from 1980 to 2016 at the extent of the urban agglomerations.
The time series of landscape metrics generally supports the diffusion and coalescence model of urban growth, with Zurich exhibiting most characteristics of coalescence while Bern and Lausanne seem to be at the transition between diffusion and coalescence.
Nevertheless, the analysis of the urban growth modes suggest that leapfrog development occurs at all periods, which contributes to an increasing fragmentation of natural patches and maintains the fractal configuration of the landscape.
The discussion reviews potential explanations for the observed landscape changes, and concludes with some planning implications.

@@

Given that a significant part of the cross-border urban agglomerations of Geneva and Basel (the second and third largest in Switzerland)) lie beyond the Swiss boundaries \cite{sfso2014espace}, in order to ensure coherence of the land use/land cover data (see the section below), this study comprises only three of the five largest Swiss urban agglomerations, namely Zurich, Bern and Lausanne \cite{sfso2018city}.
As shown in \autoref{figures/population_change.pdf}, the three cities have undergone important population growth over the last 30 years, especially during the most recent years and at the agglomeration extent.
With a total population over 1.3 million and land area of 1305 km$^2$ (1038 hab/km$^2$), Zurich is the largest Swiss urban agglomeration. As a leading global city and one of the world's largest financial centers, Zurich has the country's largest airport and railway station, and also hosts the largest Swiss universities and higher education institutions. 
% ranking among the most liveable cities in the world.
Bern is the capital of Switzerland and fourth most populous urban agglomeration in Switzerland, with a total population of 410000 inhabitants and occupying a land area of 783 km$^2$ (531 hab/km$^2$). As the fifth largest Swiss urban agglomeration and the second most important student and research center after Zurich, Lausanne has a total population of 409000 inhabitants over a land area of 773 km$^2$ (537 hab/km$^2$). Given its larger population growth rate, Lausanne is likely to soon surpass Bern and become the fourth largest urban agglomeration in Switzerland.
% Zurich: 1021859 (1990), 1080728 (2000), 1191058 (2010), 1354140 (2017)
% Bern: 351084 (1990), 349157 (1990), 357668 (2010), 415785 (2017)
% Lausanne: 300280 (1990), 311441 (2000), 339,389 (2010), 415596 (2017)
% Densities: Zurich: 1037 Bern: 531, Lausanne: 537
Overall, the three urban agglomerations are characterized by a pervasive public transportation system and a highly developed economy, with a 85\% of the employment devoted to the tertiary sector.

@@

The Swiss Federal Statistical Office (SFSO) provides an inventory of land statistics datasets \cite{sfso2017statistique}, namely a set of land use/land cover maps for the national extent of Switzerland, which comprise 72 base categories. Four datasets have been released for 1979/85, 1992/97, 2004/09 and 2013/18\footnote{\label{fn:years}The exact dates of each surveying period 1979/85, 1992/97, 2004/09 and 2013/18 are determined according to the production process of the national maps and vary accross the Swiss territory \cite{sfso2017statistique}}, at a spatial resolution of one hectare per pixel.
The pixel classification is based on computer-aided interpretation of satellite imagery, which includes special treatment and field verification of pixels where the category attribution is not clear.

The SFSO land statistics datasets have been used to produce a series of categorical maps for each urban agglomeration and time period.
The inventory is stored within a relational database where each entry corresponds to the coordinates of a given pixel and multiple columns that feature its respective land use/land cover classes for each survey period. In order to process the data in an automated and reproducible manner, an open source reusable toolbox to manage, transform and export categorical raster maps has been developed in Python \cite{bosch2019swisslandstats}.
The boundaries of each urban agglomeration have been adopted from the definitions provided also by the SFSO \cite{sfso2014espace}, which comprise multiple municipalities and have been established in consideration of population density, proximity between centers, economic activities and commuting behavior.
As stated above, Geneva and Basel are excluded from this study because a significant portion of their urban agglomeration lies beyond the extent covered by the SFSO land statistics inventory, namely the administrative boundaries of Switzerland.
% \cite{bosch2019swisslandstats}.
% The evolution of the urban agglomeration
The spatiotemporal evolution of the urban footprint for the three selected urban agglomerations (i.e., Bern, Lausanne and Zurich) over the study period (i.e., 1980-2016\textsuperscript{\ref{fn:years}}) is displayed in \autoref{figures/landscape_plots.pdf}.

@@

Firstly, the time series of landscape metrics generally support the diffusion and coalescence hypothesis, with Zurich, the largest Swiss urban agglomeration, exhibiting most characteristics of coalescence, whereas Bern and Lausanne are seemingly undergoing a transition between diffusion and coalescence. Additionally, the trend of SHDI (not explicitly considered in the diffusion and coalescence formulation of Dietzel et al. \cite{dietzel2005spatio}) suggests that the landscape is becoming increasingly diverse in land use. %, which is consistent with the findings of Wu et al. \cite{wu2011quantifying}.

@@

% While the fragmentation of the Swiss spatial planning system 
% The fragmented character of the Swiss spatial planning system is often regarded as \cite{mann2009institutional}. % , messer2017depasser}
% Planning approaches that misappreciate the residential preferences for low-density environments have often resulted 
% Ecosystem services
% The fragmentation of planning authorities is widely regarded as one of the main causes of urban sprawl \cite{carruthers2002fragmentation, mann2009institutional}, and the revision of the Swiss Federal Statue on regional planning of 2013 aims at overcoming such deficiency.
The lack of coordination between of planning authorities is widely regarded as one of the main causes of urban sprawl \cite{carruthers2002fragmentation, mann2009institutional}, and the revision of the Swiss Federal Act on regional planning of 2013 intends to overcome such shortcoming and define a national growth management strategy based on infill redevelopment and densification.
Nevertheless, growth management policies that misappreciate the residential preferences for low-density environments might inadvertently encourage households to relocate further away from the agglomeration centers, resulting in longer commute times and an overall increase of sprawl at the regional scale \cite{schwanen2004policies, robinson2005twenty}.
Additionally, the interspersion of patches of natural land within urban agglomerations provide valuable ecosystem services to its residents, such as the reduction of air pollution, alleviation of maximum temperatures, absorption of storm water, noise reduction, carbon sequestration, improvement of aesthetic and cultural values as well as the preservation of ecological habitats and biodiversity \cite{bolund1999ecosystem, gomez2013classifying}.
In Switzerland, the lowest supply of such ecosystem services is found in the cantons with highest population density, hence policies encouraging densification should consider the role of natural patches to ensure a sustainable supply of ecosystem services \cite{jaligot2019historical}.
To that end, recent planning approaches based on fractal geometry \cite{yamu2015spatial} might be exploited in order to satisfy the residential demand for green environments while explicitly considering planning objectives such as protecting natural habitats and improving accessibility to urban and natural amenities \cite{bosch2019addressing}.
% Since fractal structures have the key property that their border is over-proportionally lengthened with respect to their area.
% Land use plans should explicitly consider planning goals, fractal planning \cite{bosch2019addressing}
